---
title: "Project 2"
author: "Igor Balagula"
date: "February 17, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(recommenderlab)
library(ggplot2)
```

## Objective

Take some recommendation data and implement at least two different recommendation algorithms on the data.  For example, content-based, user-user CF, and/or item-item CF. Evaluate different approaches, using different algorithms, normalization techniques, similarity methods, neighborhood sizes, etc.

## Data

The 100K MovieLens dataset of movies will be used. The dataset contains records of 672 users and their ratings of 9066 movies. 


Load data

```{r}
setwd("E:\\Igor\\CUNY\\DATA 643 - Recommender Systems\\Projects\\Project_02\\Data")
ratings<-read.csv("ratings.csv")
movies<-read.csv("movies.csv",stringsAsFactors=FALSE)
```

Review loaded data

```{r}
head(ratings)
head(movies)
```

Remove timestap from ratings dataframe

```{r}
ratings <- ratings[,c(1,2,3)]
```



Create ratings matrix. Rows = userId, Columns = movieId

```{r}
ratingmat <- dcast(ratings, userId~movieId, value.var = "rating", na.rm=FALSE)
```

Remove userIds in the first column

```{r}
ratingmat <- as.matrix(ratingmat[,-1]) 
```

Convert rating matrix into a recommenderlab sparse matrix

```{r}
ratingmat <- as(ratingmat, "realRatingMatrix")
ratingmat
```
Let's see how ratings are distributed

```{r}
vector_ratings <- as.vector(ratingmat@data)
vector_ratings <- vector_ratings[vector_ratings != 0]
vector_ratings <- factor(vector_ratings)
qplot(vector_ratings) + ggtitle("Distribution of the ratings")
```

The most common rating is 4.

Now let's see how the average ratings for each individual movie are distributed

```{r}
average_ratings <- colMeans(ratingmat)
qplot(average_ratings) + stat_bin(binwidth = 0.1) +
ggtitle("Average movie ratings")
```

To remove some extreme cases (with very small number of ratings) we will reduce the matrix - select only those movies that were rated 50 or more times and only users who rated 50 or more movies

```{r}
ratingmat <- ratingmat[rowCounts(ratingmat)>= 50,colCounts(ratingmat)>=50 ]
ratingmat
```

Separate data into a training and test data sets (as described in the textbook p.78)

```{r}
set.seed(1)
percentage_training <-0.8
items_to_keep <- 5
rating_threshold <- 3
n_eval <- 1
eval_sets <- evaluationScheme(data=ratingmat, method="split", train=percentage_training, given=items_to_keep,
goodRating=rating_threshold,k=n_eval)
eval_sets
getData(eval_sets, "train")
```

Now we will evaluate prediction models with different parameters.

1. User-based collaborative filtering; distance method = cosine; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "Z-Score",method="Cosine",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_cosine_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_cosine_z
```

2. User-based collaborative filtering; distance method = jaccard; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "Z-Score",method="jaccard",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_jaccard_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_jaccard_z
```

3. User-based collaborative filtering; distance method = pearson; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "Z-Score",method="pearson",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_pearson_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_pearson_z
```

4. Item-based collaborative filtering; distance method = cosine; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "Z-Score",method="Cosine"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_cosine_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_cosine_z
```

5. Item-based collaborative filtering; distance method = jaccard; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "Z-Score",method="jaccard"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_jaccard_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_jaccard_z
```

6. Item-based collaborative filtering; distance method = pearson; normalize= "Z-Score"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "Z-Score",method="pearson"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_pearson_z <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_pearson_z
```

7. User-based collaborative filtering; distance method = cosine; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "center",method="Cosine",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_cosine_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_cosine_c
```

8. User-based collaborative filtering; distance method = jaccard; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "center",method="jaccard",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_jaccard_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_jaccard_c
```

9. User-based collaborative filtering; distance method = pearson; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "UBCF", param=list(normalize= "center",method="pearson",nn=30))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ubcf_pearson_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ubcf_pearson_c
```

10. Item-based collaborative filtering; distance method = cosine; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "center",method="Cosine"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_cosine_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_cosine_c
```

11. Item-based collaborative filtering; distance method = jaccard; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "center",method="jaccard"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_jaccard_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_jaccard_c
```

12. Item-based collaborative filtering; distance method = pearson; normalize= "center"

```{r}
recommender_model <- Recommender(getData(eval_sets, "train"), method = "IBCF", param=list(normalize= "center",method="pearson"))
recom <- predict(recommender_model, getData(eval_sets, "known"), type="ratings")
predict_acc_ibcf_pearson_c <- calcPredictionAccuracy(recom, getData(eval_sets, "unknown"))[1] 
predict_acc_ibcf_pearson_c
```
## Conclusion

Judging by RMSE values User-Based Collaborative Filtering provides more accurate predictions as compared to Item-Based Collaborative Filtering. Among UBCF all of the methods provide similar accuracy of predictions. However Pearson distance coupled with "center" normalization method provides the best prediction accuracy within UBCF.

